---
title: "Datathon Team 18"
author: "Merilyn Albuquerque"
date: "2025-01-21"
output:
  html_document: default
  pdf_document: default
---

#Datathon Team 18 - Script for data tidy and figures for clinical characteristics

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Merge TRACERx NSCLC dataset with MSK NSCLC

Merge and tidy up to create one neat dataframe for both studies

1. Tidy up MSK data
2. Tidy up metastatic sites from multiple columns into one column ('Metastatic.Site') in Supplementary_Table14 document
3. Add corresponding columns to TRACERx data to match MSK (i.e. 'Study.ID', 'Cancer.Type', 'Histology')
4. Subset relevant columns in TRACERx, rename column names to match MSK
5. Merge TRACERx with MSK data, save file
6. Plot tables and graphs of each dataset separately and together

Things to note:

1. MSK data has duplicate patient IDs for each sample ID (multiple samples per patient) - this needs to be addressed when looking at clinical characteristics and when performing regression modelling
2. Conversely, TRACERx data has multiple met sites for one patient ID - but this can be easily tidied for better plotting
3. Overall survival in TRACERx data did not mention if it was in days (we assumed and calculated time in months to match MSK data)
4. Smoking status in MSK is TRUE (Smoker and Ex-Smoker) and FALSE (Never Smoked). Labels were changed in both datasets to help group patients into these two categories.


```{r data}
library(dplyr)
library(tidyr)

#Load all data - TRACERx dataset, MSK dataset, Supplementary Table 14 file
all_patients_df <- readRDS("figurecode/data/20221109_TRACERx421_all_patient_df.rds")
msk_table_raw <- read.csv("MSK_NSCLC_2022_raw.csv")
supp_t14 <- readxl::read_excel("Supp_Table_14.xlsx")


#Tidy data in MSK 
msk_table <- msk_table_raw[,c("Study.ID", "Patient.ID", "Sample.Type", "Cancer.Type", "Cancer.Type.Detailed", 
                                  "Oncotree.Code", "Overall.Survival..Months.", "Sex", "Smoking.Status", "Patient.Current.Age", 
                                  "Ethnicity.Category", "Histology", "Metastatic.Site")]
msk_table <- msk_table %>% mutate(Metastatic.Site = ifelse(Metastatic.Site == "Lymph node", "Lymph Node", Metastatic.Site))
msk_table <- msk_table %>% mutate(Smoking.Status = ifelse(Smoking.Status == "True", "Smoker/Ex-Smoker", "Never Smoked"))
colnames(msk_table) <- c("Study.ID", "Patient.ID", "Sample.Type", "Cancer.Type", "Cancer.Type.Detailed", 
                                  "Oncotree.Code", "Overall.Survival..Months.", "Sex", "Smoking.Status", "Age", 
                                  "Ethnicity", "Histology", "Metastatic.Site")

#Adjust met site info from Supplementary Table 14 file, merge with TRACERx file
supp_t14 <- supp_t14[-c(1:4),-c(2,4)]
colnames(supp_t14) <- supp_t14[1,]
supp_t14 <- supp_t14[-1,]
names(supp_t14)[names(supp_t14) == "PublicationID"] <- "cruk_id"
supp_t14 <- supp_t14 %>%
  mutate(across(-cruk_id, as.character))
supp_t14 <- supp_t14 %>%
  pivot_longer(-c(cruk_id), names_to = "Condition", values_to = "Value") %>% 
  filter(Value == "TRUE") %>%  
  group_by(cruk_id) %>%
  summarize(Combined_Conditions = paste(Condition, collapse = ", "), .groups = "drop")
colnames(supp_t14) <- c("cruk_id", "Metastatic.Site")
tracex_met_pts_df <- left_join(all_patients_df, supp_t14, by = "cruk_id")

#Tidy data in TRACERx (general tidy and adding columns to match MSK data for better merged df)
tracex_met_pts_df <- tracex_met_pts_df %>% 
  mutate(Relapse_cat_new = case_when(Relapse_cat_new == "No rec" ~ "No metastasis", TRUE ~ as.character(Relapse_cat_new)))
tracex_met_pts_df <- tracex_met_pts_df %>% 
  mutate(os_time = os_time/30.44)
tracex_met_pts_df$Study.ID <- "nsclc_TRACERx_2023"
tracex_met_pts_df$Cancer.Type <- "Non-Small Cell Lung Cancer"
tracex_met_pts_df$histology_multi_full_genomically.confirmed <- gsub("x2|x3", "", tracex_met_pts_df$histology_multi_full_genomically.confirmed)
tracex_met_pts_df <- tracex_met_pts_df %>% 
  mutate(Histology = case_when(grepl("LUAD", histology_multi_full_genomically.confirmed) ~ "Adenocarcinoma",
                               grepl("LUSC", histology_multi_full_genomically.confirmed) ~ "Squamous Cell Carcinoma",
                               TRUE ~ histology_multi_full_genomically.confirmed))
tracex_met_pts_df <- tracex_met_pts_df %>%
  mutate(Sample.Type = case_when(grepl("Intrathoracic", Relapse_cat_new) ~ "Metastasis",
                                 grepl("Extrathoracic", Relapse_cat_new) ~ "Metastasis",
                                 grepl("Intra & Extra", Relapse_cat_new) ~ "Metastasis",
                                 grepl("Unknown Site", Relapse_cat_new) ~ "Unknown",
                                 TRUE ~ Relapse_cat_new))
tracex_met_pts_df <- tracex_met_pts_df %>% 
  mutate(smoking_status_merged = case_when(grepl("Smoker", smoking_status_merged) ~ "Smoker/Ex-Smoker",
                                    grepl("Ex-Smoker", smoking_status_merged) ~ "Smoker/Ex-Smoker",
                                    TRUE ~ smoking_status_merged))


tracex_met_pts_df <- tracex_met_pts_df[,c("Study.ID", "Cancer.Type", "cruk_id", 
                                          "age", "sex", "ethnicity",
                                          "os_time", "smoking_status_merged", "histology_multi_full_genomically.confirmed", 
                                          "Histology", "Relapse_cat_new", "Metastatic.Site", "Sample.Type")]
colnames(tracex_met_pts_df) <- c("Study.ID", "Cancer.Type", "Patient.ID", 
                                 "Age", "Sex", "Ethnicity", 
                                 "Overall.Survival..Months.", "Smoking.Status", "Oncotree.Code", 
                                 "Histology", "Metastatic.Category", "Metastatic.Site", "Sample.Type")

#Merge MSK dataset with TRACERx
MSK_TRACERx_merged <- full_join(tracex_met_pts_df, msk_table, by = c("Study.ID", "Patient.ID", "Cancer.Type", 
                                                                     "Oncotree.Code", "Overall.Survival..Months.", "Sex", 
                                                                     "Smoking.Status", "Histology", "Metastatic.Site",
                                                                     "Sample.Type", "Age"))
head(MSK_TRACERx_merged, 5)
tail(MSK_TRACERx_merged, 5)

unique_patients <- unique(MSK_TRACERx_merged$Patient.ID) #there are 1548 unique patient IDs

#write.csv(MSK_TRACERx_merged, "MSK_TRACERx_merged.csv", row.names = FALSE)

```

## TRACERx Clinical Characteristics

```{r table.tracerx}
library(tableone)

myVars <- c("Age", "Sex", "Ethnicity", "Overall.Survival..Months.", "Smoking.Status", "Histology", "Metastatic.Category")
catVars <- c("Sex", "Ethnicity", "Smoking.Status", "Histology", "Metastatic.Category")
table1 <- CreateTableOne(vars = myVars, data = tracex_met_pts_df, factorVars = catVars)
table1

```

```{r graph.tracerx}
library(ggplot2)

#Calculate percentages for each met site
pie_data_tracerx <- tracex_met_pts_df %>%
  filter(!is.na(Metastatic.Site)) %>% 
  count(Metastatic.Site, name = "Count")  

#Separate multiple met sites into individual sites for better graph
pie_data_tracerx <- pie_data_tracerx %>%
  mutate(Metastatic.Site = strsplit(Metastatic.Site, ", ")) %>% 
  unnest(Metastatic.Site) %>%                                 
  group_by(Metastatic.Site) %>% 
  summarise(Count = sum(Count)) %>% 
  mutate(Percentage = Count / sum(Count) * 100) %>% 
  arrange(desc(Percentage))


#Plot pie chart
ggplot(pie_data_tracerx, aes(x = "", y = Percentage, fill = Metastatic.Site)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y") + 
  theme_void() + 
  labs(title = "Distribution of Metastatic Sites in TRACERx Data", fill = "Metastatic Site") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_brewer(palette = "Paired") + 
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 4)


```

## MSK Clinical Characteristics

```{r table.msk}
myVars2 <- c("Age", "Sex", "Ethnicity", "Overall.Survival..Months.", "Smoking.Status", "Histology", "Sample.Type")
catVars2 <- c("Sex", "Ethnicity", "Smoking.Status", "Histology", "Sample.Type")
table2 <- CreateTableOne(vars = myVars2, data = msk_table, factorVars = catVars2)
table2

```

```{r graph.msk}
library(RColorBrewer)

#Calculate percentages for each met site
pie_data_msk <- msk_table %>%
  filter(!is.na(Metastatic.Site)) %>% 
  count(Metastatic.Site, name = "Count") %>% 
  mutate(Percentage = Count / sum(Count) * 100) %>% 
  arrange(desc(Percentage))

#Plot pie chart
#ggplot(pie_data_msk, aes(x = "", y = Percentage, fill = Metastatic.Site)) +
#  geom_bar(stat = "identity", width = 1) + 
#  coord_polar("y") + 
#  theme(plot.title = element_text(hjust = 0.5)) +
#  theme_void() + 
#  labs(title = "Distribution of Metastatic Sites in MSK Data", fill = "Metastatic Site") 


#Too many met sites - plot top 20 with remaining labelled as "Other"
top_20_msk <- pie_data_msk %>%
  arrange(desc(Percentage)) %>%
  mutate(Metastatic.Site = ifelse(row_number() > 20, "Other", Metastatic.Site)) %>%
  group_by(Metastatic.Site) %>%
  summarise(Count = sum(Count), Percentage = sum(Percentage)) 

set3_colors <- brewer.pal(12, "Set3")
additional_colours <- c("#FF6347", "#20B2AA", "#FFD700", "#8A2BE2", "#FF4500", "#2E8B57", "#DA70D6", "#F4A460", "#D2691E")
colours <- c(set3_colors, additional_colours)

ggplot(top_20_msk, aes(x = "", y = Percentage, fill = Metastatic.Site)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y") + 
  theme_void() + 
  labs(title = "Distribution of Metastatic Sites in MSK Data", fill = "Metastatic Site") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = colours) +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 4)


```



## Merged (TRACERx & MSK) Clinical Characteristics

```{r table.merged}
#Initial table for all sample types we have across entire cohort
myVar3.1 <- c("Sample.Type")
catVars3.1 <- c("Sample.Type")
table3.1 <- CreateTableOne(vars = myVar3.1, data = MSK_TRACERx_merged, factorVars = catVars3.1)
table3.1 #3042 samples in total across 1548 patients

#Need to adjust for duplicate patient rows in MSK data to get info for remaining clinical columns
msk_table_no_pts_dups <- msk_table
msk_table_no_pts_dups[is.na(msk_table_no_pts_dups)] <- "Unknown"
msk_table_no_pts_dups <- distinct(msk_table, Patient.ID, .keep_all = TRUE)
msk_table_no_pts_dups$Age <- as.integer(msk_table_no_pts_dups$Age)

MSK_TRACERx_merged_no_pts_dups <- full_join(tracex_met_pts_df, msk_table_no_pts_dups, by = c("Study.ID", "Patient.ID", "Cancer.Type", 
                                                                     "Oncotree.Code", "Overall.Survival..Months.", "Sex", 
                                                                     "Smoking.Status", "Histology", "Metastatic.Site",
                                                                     "Sample.Type", "Age"))
unique_patients <- unique(MSK_TRACERx_merged_no_pts_dups$Patient.ID) #there are 1548 unique patient IDs

myVars3.2 <- c("Age", "Sex", "Ethnicity", "Overall.Survival..Months.", "Smoking.Status", "Histology")
catVars3.2 <- c("Sex", "Ethnicity", "Smoking.Status", "Histology")
table3.2 <- CreateTableOne(vars = myVars3.2, data = MSK_TRACERx_merged_no_pts_dups, factorVars = catVars3.2)
table3.2


```

```{r graph.merged}

#Calculate percentages for each met site
pie_data_merged <- MSK_TRACERx_merged %>%
  filter(!is.na(Metastatic.Site)) %>% 
  count(Metastatic.Site, name = "Count")  

#Separate multiple met sites for TRACERx into individual sites for better graph
pie_data_merged <- pie_data_merged %>%
  mutate(Metastatic.Site = strsplit(Metastatic.Site, ", ")) %>% 
  unnest(Metastatic.Site) %>%                                 
  group_by(Metastatic.Site) %>% 
  summarise(Count = sum(Count)) %>% 
  mutate(Percentage = Count / sum(Count) * 100) %>% 
  arrange(desc(Percentage))

#Too many met sites - plot top 20 with remaining labelled as "Other"
top_20_merged <- pie_data_merged %>%
  arrange(desc(Percentage)) %>%
  mutate(Metastatic.Site = ifelse(row_number() > 20, "Other", Metastatic.Site)) %>%
  group_by(Metastatic.Site) %>%
  summarise(Count = sum(Count), Percentage = sum(Percentage)) %>% ungroup


set3_colors <- brewer.pal(12, "Set3")
additional_colours <- c("#FF6347", "#20B2AA", "#FFD700", "#8A2BE2", "#FF4500", "#2E8B57", "#DA70D6", "#F4A460", "#D2691E")
colours <- c(set3_colors, additional_colours)

ggplot(top_20_merged, aes(x = "", y = Percentage, fill = Metastatic.Site)) +
  geom_bar(stat = "identity", width = 1) + 
  coord_polar("y") + 
  theme_void() + 
  labs(title = "Distribution of Metastatic Sites in Merged Data", fill = "Metastatic Site") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_fill_manual(values = colours, guide = guide_legend(ncol = 2)) +
  geom_text(aes(label = paste0(round(Percentage, 0), "%")), 
            position = position_stack(vjust = 0.5), 
            color = "black", size = 4)


```


```{r survival}
library(survival)
library(survminer)
library(patchwork)

surv_obj <- Surv(time = MSK_TRACERx_merged$Overall.Survival..Months.)
fit <- survfit(surv_obj ~ 1, data = MSK_TRACERx_merged)
p1 <- ggsurvplot(
  fit,
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curve in Merged Dataset",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  legend = "none")

fit2 <- survfit(Surv(Overall.Survival..Months.) ~ Metastatic.Category, data = MSK_TRACERx_merged)
p2 <- ggsurvplot(
  fit2,
  data = MSK_TRACERx_merged,
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curve by Metastatic Category in TRACERx Data",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  legend = "right")

fit3 <- survfit(Surv(Overall.Survival..Months.) ~ Sample.Type, data = MSK_TRACERx_merged) 
p3 <- ggsurvplot(
  fit3,
  data = MSK_TRACERx_merged,
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curve by Sample Type in Merged Data",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  legend = "right")

fit4 <- survfit(Surv(Overall.Survival..Months.) ~ Smoking.Status, data = MSK_TRACERx_merged)
p4 <- ggsurvplot(
  fit4,
  data = MSK_TRACERx_merged,
  ggtheme = theme_minimal(),
  title = "Kaplan-Meier Curve by Smoking Status in Merged Data",
  xlab = "Time (Months)",
  ylab = "Survival Probability",
  legend = "right")

p1$plot
p2$plot
p3$plot
p4$plot

#ggsave("KP_plot_merged_data.png", plot=last_plot(), width = 8, height = 5)

```


